---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Homework 3A"
author: "Insert Your Name Here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
---

#Load your relevant libraries
```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(foreign)
library(epiDisplay)
library(stats)
library(Hmisc)
library(gmodels)

```

# Problem 1
## Create vectors called 'ht' and 'wt' corresponding to the heights and weights of the 10 family members.
## Assign the family members to the 'names' attribute of these vectors.
## Make a data frame with ht and wt.
## Create a new variable in the data frame, body mass index (BMI) of each person where BMI = wt/ht2.
## Print the data frame in your R output or take a screen shot of your data frame.
## Save the data frame as a permanent R data object.

```{r problem 1}
# Insert code for problem 1 here
```
# Interpret your answer for problem 1 here


# Problem 2
## Export the data frame as a csv file. Attach a screen shot of the file.
```{r problem 2}
# Insert code for problem 2 here
```
# Interpret your answer for problem 2 here


# Problem 3
## Identify the persons who have the lowest and highest BMI.
```{r problem 3}
# Insert code for problem 3 here
```
# Interpret your answer for problem 3 here


# Problem 4
## Calculate the mean, median and SD of BMI.
```{r problem 4}
# Insert code for problem 4 here
```
# Interpret your answer for problem 4 here

