---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Data Management with R"
author: "Kelly Bakulski"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
   word_document: default
   html_document: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # This is how you set options for all chunks of code
```


## Chapter 3, Data Management with R
# Install new packages
```{r install packages, eval=FALSE}

# Install packages. Do this only once.
options(repos="https://cran.rstudio.com" )
install.packages("foreign")
install.packages("Hmisc")
install.packages("stargazer")
install.packages("compareGroups")
install.packages("car")

# To avoid installing every time: change set up in curly brackets to eval=FALSE
```


# Set up: Query the current R environment and load relevant packages
```{r packages}
search() # list the packages, environments, or data frames
ls() # list the objects

##### Load these packages for the current session
library(foreign)
library(Hmisc)
library(epiDisplay)
library(stargazer)
library(compareGroups)
library(car)

sessionInfo() # record the versions of packages used in the current session
```

# Specify file directories
```{r directories}
directory <- "/cloud/project/" #Class option when coding on RStudio Cloud

# Three alternate options to customize and specify source data directory if programming on a personal computer (only need to do the one that works for you)
# directory<-"M:/EPID674/Data/" # Alternate option 1: location of the parent dataset in afs space of school computer
#directory <- "/Volumes/GoogleDrive/My Drive/Teaching/EPID674/2020_fall/EPID674_Week3_Class/" # Alternate option 2: location of data on my Macbook
# directory<-"C:/Users/bakulski/Documents/EPID674/Data/" #Alternate option 3: location of data on a personal PC

date <- format(Sys.Date(), "%Y%m%d")
```


# Load data, attach data, remake useful variables
```{r read data}

# Check the file path
paste0(directory, "nhanes3.rda")
# Load the saved R data
load(paste0(directory, "nhanes3.rda"))
### Or load RData if you saved the workspace image
# load(".RData")


# Remake a few variables from last class if they are no longer in your environment
sex1 <- factor(nhanes$sex, levels = c(1, 2), labels = c("male", "female"))
AGE5b <- cut(nhanes$age, quantile(nhanes$age, c(0, .2, .4, .6, .8, 1)), include.lowest = T) # quintiles
AGE5c <- cut(nhanes$age, breaks = c(19, 40, 50, 60, 70, 90))
age5c <- unclass(AGE5c)
```

# Assign names to objects, bind columns to dataframe
```{r names and cbind}
# Assign the subject IDs as the rownames
rownames(nhanes)[1:10]
nhanes$seqn[1:10]
rownames(nhanes) <- nhanes$seqn
rownames(nhanes)[1:10]

### Bind vectors/matrices/data
nhanes.a <- cbind(nhanes, sex1, AGE5b, age5c) # bind by column (merge)
head(nhanes.a)
```

# Set up a simulated dataset with unequal rows and columns, attempt to bind
```{r simulate uneven dataset}
nhanes.left <- nhanes[, 1:20]
nhanes.right <- nhanes[, c(2, 21:37)]
names(nhanes.left)
names(nhanes.right)

nhanes.right <- nhanes.right[-c(1:30), ]
dim(nhanes.left)
dim(nhanes.right)
# nhanes.b<-cbind(nhanes.left, nhanes.right)
# does not work b/c of different number of rows, must comment out for knitting

## let's also shuffle nhanes3.right
nhanes.right1 <- nhanes.right[sample(nrow(nhanes.right)), ]
nhanes.right[1:10, 1:3]
nhanes.right1[1:10, 1:3]
```

# Use the merge() function to combine datasets, explore function options
```{r merge}
nhanes.both <- merge(nhanes.left, nhanes.right1, by = "seqn")
names(nhanes.both)
nhanes[1:40, c(1:3, 21:22)]
nhanes.both[1:40, c(1:3, 21:22)]
dim(nhanes)
dim(nhanes.both)

nhanes.both1 <- merge(nhanes.left, nhanes.right1, by = "seqn", all.x = T)
dim(nhanes.both1)
nhanes.both1[1:40, c(1:3, 21:22)]

nhanes.both2 <- merge(nhanes.right1, nhanes.left, by = "seqn", all.x = T)
dim(nhanes.both2)
nhanes.both2[1:40, c(1:3, 21:22)]

nhanes.both3 <- merge(nhanes.right1, nhanes.left, by = "seqn", all.y = T)
dim(nhanes.both3)
nhanes.both3[1:40, c(1:3, 21:22)]
```

# Sorting/ordering data
```{r sort order}
nhanes$age[1:10]
sort(nhanes$age)[1:10]
sort(nhanes$age, decreasing = TRUE)[1:10] # reverse sort

nhanes[1:10, c("seqn", "age", "sex")]
nhanes.b <- nhanes[order(nhanes$age), ]
nhanes.b[1:10, c("seqn", "age", "sex")]
nhanes.c <- nhanes[order(nhanes$age, nhanes$sex), ] # sort based on 2 variables
nhanes.c[1:10, c("seqn", "age", "sex")]
```

# Create a new variable in a dataset using calculations
```{r create numeric variable}
## Create a new variable in a dataset using calculations
nhanes$map <- nhanes$dbp + (nhanes$sbp - nhanes$dbp) / 3
summary(nhanes$map)

nhanes.b$map <- with(nhanes.b, dbp + (sbp - dbp) / 3)
summary(nhanes.b$map)

nhanes.c <- transform(nhanes.c, map = dbp + (sbp - dbp) / 3)
summary(nhanes.c$map)
```


# Recode a variable or replace a value within a variable
```{r recode replace}
# Replace
nhanes$age1 <- replace(nhanes$age, nhanes$age == 90, NA)
summary(nhanes$age1)

# Recode
# the recode function in the library 'car'

nhanes[c(1:20), ]
# convert sex=2 to 0
nhanes$male <- recode(nhanes$sex, "2=0; else=nhanes$sex")
nhanes[c(1:20), c("sex", "male")]

# Recode 65+ as 1, other as 2
nhanes$age2 <- recode(nhanes$age, "65:90=1; else=2")
nhanes[c(1:20), c("age", "age2")]

### please note that the function 'recode' in epiDisplay is different than in the car package.
### which is similar to the replace() function
### use package::function to overcome masking
```

# Do Exercise 3A

# Check function default settings for handling missing values
```{r check default missing}
### How do functions handle missing values?
cor(nhanes$age, nhanes$bmi)
summary(nhanes$bmi)
summ(nhanes$bmi)
# What options can you use to overcome the default missing data setting?
cor(nhanes$age, nhanes$bmi, use = "complete")
mean(nhanes$bmi)
mean(nhanes$bmi, na.rm = T)
mean(na.omit(nhanes$bmi))
```

# Use the na.omit() function to drop missing observations, make a new dataset
```{r na.omit}
nhanes.d <- na.omit(nhanes[, c(
  "age", "bpb", "sbp", "dbp", "bmi", "packyrs",
  "hematoc", "educ", "smk", "alc", "med_ht", "diag_dm"
)])
nhanes.e <- na.omit(nhanes)
dim(nhanes.d)
dim(nhanes.e)

cor(nhanes.d$age, nhanes.d$bmi)
cor(nhanes.e$age, nhanes.e$bmi)
```

# Use a logical vector and square brackets to index non-missing observations
```{r logical index}
OK <- !is.na(nhanes$age) & !is.na(nhanes$bmi)
cor(nhanes$age[OK], nhanes$bmi[OK])

OK <- OK & !is.na(nhanes$hematoc) & !is.na(nhanes$chol) & !is.na(nhanes$educ) & !is.na(nhanes$smk) & !is.na(nhanes$packyrs) &
  !is.na(nhanes$alc) & !is.na(nhanes$med_ht) & !is.na(nhanes$diag_dm)
cor(nhanes$age[OK], nhanes$bmi[OK])
cor(nhanes$age[OK == 1], nhanes$bmi[OK == 1])
```

# Strategies to make repeated tasks in your coding more efficient

# Use the apply family of functions for more efficient coding
```{r apply}
sapply(nhanes, mean)
tapply(nhanes$bmi, age5c, mean)
tapply(nhanes$bmi, age5c, summary)

sapply(nhanes, mean, na.rm = T)
# you can get the same results using apply() or lapply().
apply(nhanes, 2, mean, na.rm = T)
# 2 indicates columns, 1 does rows
lapply(nhanes, mean, na.rm = T)

codebook(nhanes.a) # returns summary stat by variable characteristics
summ(nhanes)
```

# Try compiling results with a for loop. 
```{r for loop}
# First prep the output dataset
out <- data.frame(matrix(nrow = ncol(nhanes), ncol = 3))
colnames(out) <- c("variable", "mean", "sd")
out[, 1] <- colnames(nhanes)
# Then initiate the loop
for (i in 1:ncol(nhanes)) {
  out[i, 2] <- mean(as.numeric(nhanes[, i]), na.rm = T)
  out[i, 3] <- sd(as.numeric(nhanes[, i]), na.rm = T)
}
out
```


# Try compiling/exporting results with stargazer
```{r stargazer}
stargazer(nhanes)
stargazer(nhanes, type = "text", title = "Descriptive Statistics", digits = 1)
# stargazer(nhanes, type="text", title="Descriptive Statistics", digits=1, out="Descriptive.txt")
```


# Generate descriptive statistics for each group
```{r compareGroups}
sapply(nhanes[nhanes$age >= 65, ], mean, na.rm = T) # for a subset age>=65
tapply(nhanes$bmi, age5c, mean, na.rm = T) # stratified by age

# Try compiling/exporting results by group with compareGroups
createTable(compareGroups(nhanes)) # for all variables
createTable(compareGroups(sex1 ~ . - strata - seqn, data = nhanes.a)) # for all variables, stratified by sex

## similar grouped outputs can be obtained using aggregate()
aggregate(nhanes$bmi, by = list(Sex = nhanes$sex, Race = nhanes$race), mean, na.rm = T) # stratified by sex and race
aggregate(nhanes[, c("sbp", "dbp")], by = list(Sex = nhanes$sex, Race = nhanes$race), mean) # for multiple columns at once
```


# Date and Time 

# Input variable formats, convert to desired format
```{r date format}
# create some dates to use in this exercise
bdays <- c("11/2/1959", "1/1/1970")
bdays
# convert to Julian dates
bdays.julian <- as.Date(bdays, format = "%m/%d/%Y")
bdays.julian

as.numeric(bdays.julian)

bdays.am <- format(bdays.julian, "%b %d, %Y")
bdays.am

# day of the week
format(bdays.julian, "%a-%d%b%y")
weekdays(bdays.julian)
```


# Calculate age as of today's date
```{r age from date}
date.today <- Sys.Date() # Sys.time or Sys.Date: Current time/Date
date.today

agecomp <- (date.today - bdays.julian) / 365.25
agecomp

agecomp2 <- trunc(as.numeric(agecomp))


# create data frame
bd <- data.frame(Birthday = bdays, Standard = bdays.julian, Julian = as.numeric(bdays.julian), Age = agecomp2)
bd
```


# Paste function
```{r paste}
# combine (paste) two or more variables that are parts of date
day1 <- c("12", "13", "14")
month1 <- c("07", "08", "12")
year1 <- c("1980", "2000", "2010")
paste(day1, month1, year1)

as.Date(paste(day1, month1, year1), "%d %m %Y")
```

# Do Exercise 3B

