---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Descriptive Statistics, Tables, Misc."
author: "Kelly Bakulski, Lauren Middleton"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # This is how you set options for all chunks of code
```

## Chapter 3, Descriptive Statistics, Tables, Misc.

# Install new packages

```{r install packages, eval=FALSE}

# Install packages. Do this only once.
# We have this completed for you on Cloud, so you do not need to run this.
options(repos="https://cran.rstudio.com" )
install.packages("tidyverse")
install.packages("here")
install.packages("gtsummary")
install.packages("flextable")

# To avoid installing every time: change set up in curly brackets to eval=FALSE
```

# Set up: Query the current R environment, load relevant packages, load dataset created in Week 2

```{r packages}

search() # list the packages, environments, or data frames
ls() # list the objects

##### Load these packages for the current session
library(tidyverse)
library(here)
library(gtsummary)
library(flextable)

sessionInfo() # record the versions of packages used in the current session

# Load saved NHANES dataset from Week 2
load(here(("nhanes_dataset.rda")))
```


# Data description functions, numeric variables

```{r univariate_descriptives_numeric_variable}
# Base R method - quick, good for one variable at a time
summary(nhanes_dataset$RIDAGEYR)
sd(nhanes_dataset$RIDAGEYR)

# Tidyverse method - more code, but can be used to get summary statistics on multiple variables
nhanes_dataset %>%
  summarise(age_min = min(RIDAGEYR),
            age_25_quart = quantile(RIDAGEYR, probs = 0.25),
            age_mean = mean(RIDAGEYR),
            age_median = median(RIDAGEYR),
            age_75_quart = quantile(RIDAGEYR, probs = 0.75),
            age_iqr = IQR(RIDAGEYR),
            age_max = max(RIDAGEYR))

```


# Check your understanding! 
Calculate descriptive statistics on the poverty income variable: INDFMPIR
Specifically, calculate the minimum, 25th percentile, median, mean, 75th percentile, and maximum. 
* Start by creating a new r code chunk
* Pick your method for calculating descriptive statistics!



# Calculate descriptive statistics on a categorical variable

```{r categorical_variable_descriptives}

# Base R method- quick, good for one variable at a time
table(nhanes_dataset$race_eth)

# Tidyverse method - more code, but can be used to get summary statistics on combinations of multiple variables
nhanes_dataset %>% 
	count(race_eth) %>%
	mutate(percent = n / sum(n) * 100)

```
# Reproducible, publication quality univariate tables

```{r reproducible_univariate_tables}

# Univariate table using gtsummary package
colnames(nhanes_dataset)
nhanes_dataset %>%
  select(RIDAGEYR, 
         sex, 
         race_eth,
         education,
         LBXRBCSI,
         LBXWBCSI,
         LBDLYMNO,
         LBDNENO,
         nlr,
         LBXIRN,
         iron_status,
         URXUAS,
         LBXBCD,
         LBXBPB,
         LBXCOT) %>% #drop SEQN and survey variables so we don't get summary statistics for the identifiers or weights
  tbl_summary(label = list(education ~ "Overall Educational Attainment", 
                           race_eth ~ "Race/ethnicity"),
             statistic = all_continuous() ~ "{mean} ({sd})")
```



### Filter, order, bivariate descriptive statistics

# Ordering data

```{r order_rows}

# Look at the current order of rows
head(nhanes_dataset)

# Order by age and then sex
nhanes_dataset %>%
  arrange(RIDAGEYR, sex) %>%
  head()

# Descending order by age and then sex
nhanes_dataset %>%
  arrange(desc(RIDAGEYR, sex)) %>%
  head()

```

# Filter to subset the dataset

```{r filter_data}

# Practice filtering: Keep only the female participants
table(nhanes_dataset$sex) # What do you expect?
nhanes_subset <- nhanes_dataset %>%
  filter(sex == "Female")
dim(nhanes_subset) # What do you get?

# Keep only the participants who are iron deficient and older than 60 years using filter()
table(nhanes_dataset$iron_status, nhanes_dataset$RIDAGEYR >= 60) # What do you expect?
nhanes_subset <- nhanes_dataset %>%
  filter(iron_status == "Deficient" & RIDAGEYR >= 60)
dim(nhanes_subset) # What do you get?

# Now can calculate univariate descriptive statistics on the subset datasets

```

# Calculate bivariate descriptive statistics

```{r bivariate_descriptives_numeric_variable}
# Calculating statistics by groups - tidyverse
# Note: good practice to add ungroup() after using it to avoid the dataset staying grouped
nhanes_dataset %>%
  group_by(sex) %>%
  summarise(age_min = min(RIDAGEYR),
            age_25_quart = quantile(RIDAGEYR, probs = 0.25),
            age_mean = mean(RIDAGEYR),
            age_median = median(RIDAGEYR),
            age_75_quart = quantile(RIDAGEYR, probs = 0.75),
            age_iqr = IQR(RIDAGEYR),
            age_max = max(RIDAGEYR)) %>%
  ungroup()

```

# Calculate bivariate descriptive statistics on categorical variables

```{r bivariate_descriptives_categorical_variable}

# Base R method
table(nhanes_dataset$race_eth, nhanes_dataset$sex)

# Tidyverse method
nhanes_dataset %>% 
	count(race_eth, sex) %>%
	mutate(percent = n / sum(n) * 100)

```

# Check your understanding!

Calculate the bivariate descriptive statistics between the iron status and age group variables: iron_status and cut_groups
* Start by creating a new r code chunk
* Identify the types of variables to determine the descriptive statistics to calculate
* Try calculating them! 



## Bivariate descriptive statistics, publication ready table

```{r bivariate_descriptive_table}

# Bivariate table by sex using gtsummary - will take about 5 seconds to run
bivar_nhanes <- nhanes_dataset %>%
  select(RIDAGEYR, 
         sex, 
         race_eth,
         education,
         INDFMPIR,
         LBXRBCSI,
         LBXWBCSI,
         LBDLYMNO,
         LBDNENO,
         nlr,
         LBXIRN,
         iron_status,
         URXUAS,
         LBXBCD,
         LBXBPB,
         LBXCOT) %>%
  tbl_summary(by = sex, #stratify by sex
              label = list(RIDAGEYR    ~ "Age (years)", #update the variable names
                           race_eth    ~ "Race/Ethnicity",
                           INDFMPIR    ~ "Poverty-Income Ratio",
                           LBXRBCSI    ~ "Red Blood Cell Count (million cells/uL)",
                           LBXIRN      ~ "Iron (ug/dL)",
                           iron_status ~ "Iron Status",
                           nlr         ~ "Neutrophil:Lymphocyte Ratio"),
              statistic = list(all_continuous() ~ "{mean} ({sd})", #use mean and standard deviation for continuous variables
                               all_categorical() ~ "{n} ({p}%)"),  #use count and percentage for categorical variables
              digits = list(all_categorical() ~ c(0, 1), #adds no decimal places to counts and one decimal to percentages
                            all_continuous() ~ 1), #adds one decimal place to mean and sd values
              missing_text = "Missing (n)"
              ) %>%
  add_p() %>% #compares male vs female, does not include overall in p-value calculation
  add_overall() %>% #adds column with non-stratified summary statistics
  modify_header(label ~ "**Variable**") %>% #the asterisks bold the label
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Sex**") %>%
  bold_labels()

#View table
bivar_nhanes


# Save the table as a Word document - some formatting such as bold headers will be lost
bivar_nhanes %>%
  as_flex_table() %>%
  save_as_docx(path = here("summary_stats_table.docx")) #Export the table to Word using flextable package
```


# Missing data, date variables, for loops

# Do any of our variables have missing values?
```{r missing_data}

# How many participants have complete data on all variables?
nhanes_dataset %>%
  na.omit() %>% #exclude any participants with missing data
  dim() #what are the dimensions after excluding missing data?

# What variables have the most missing data?
summary(nhanes_dataset)

```

# Check function default settings for handling missing values
```{r check_default_missing}
### How do functions handle missing values?

#calculate summary statistics on iron
summary(nhanes_dataset$LBXIRN)

#calculate the mean of iron
mean(nhanes_dataset$LBXIRN)
mean(nhanes_dataset$LBXIRN, na.rm = TRUE)
#note the na.omit - remove all rows that include NAs
mean(na.omit(nhanes_dataset$LBXIRN))

#calculate a correlation between age and iron concentration
cor(nhanes_dataset$RIDAGEYR, nhanes_dataset$LBXIRN)
cor(nhanes_dataset$RIDAGEYR, nhanes_dataset$LBXIRN, use="complete.obs")

```


# Try compiling results with a for loop

```{r for_loop}

# Filter to the numeric variables, we'll be calculating numeric descriptive statistics
nhanes_numeric <- nhanes_dataset %>%
    select_if(is.numeric)

# First prep the output dataset - set up blank columns
out <- data.frame(matrix(nrow = ncol(nhanes_numeric), ncol = 3))

# Set the column names for the dataframe
colnames(out) <- c("variable", "mean", "sd")

# Pull the column names from nhanes_dataset and set them as the values for the first column
out[, 1] <- colnames(nhanes_numeric)

# Then initiate the loop
for (i in 1:ncol(nhanes_numeric)) {
  out[i, 2] <- round(mean(as.numeric(nhanes_numeric[, i]), na.rm = TRUE), digits = 2)
  out[i, 3] <- round(sd(as.numeric(nhanes_numeric[, i]), na.rm = TRUE), digits = 2)
}
out

```

# Date and Time objects


# Paste function

```{r paste}

# create some dates to use in this exercise
day <- c("1", "12", "13", "2")
month <- c("1", "07", "08", "11")
year <- c("1970","1980", "2000", "1959")

# combine (paste) two or more variables that are parts of date
birth_days_chr <- paste(month, day, year, sep = "/")
str(birth_days_chr)
```

# Input date information as character, convert to date format

```{r date_format}

# convert to dates
birth_days_date <- as.Date(birth_days_chr, format = "%m/%d/%Y")
str(birth_days_date)

as.numeric(birth_days_date)

birth_days_month <- format(birth_days_date, "%b %d, %Y")
birth_days_month

# day of the week
format(birth_days_date, "%a-%d%b%y")
weekdays(birth_days_date)
```

# Calculate age as of today's date

```{r age_from_date}

date_today <- Sys.Date() # Sys.time or Sys.Date: Current time/Date
date_today

current_age <- (date_today - birth_days_date) / 365.25
current_age

current_age <- trunc(as.numeric(current_age))


# create data frame
bd <- data.frame(Character = birth_days_chr, Standard_Date = birth_days_date, Numeric_Date = as.numeric(birth_days_date), Age = current_age)
bd
```

# Remember to save your R script!

# To exit R

```{r exit, eval=F}
# q()
## if you close R, you will be asked to save your workspace image
```