---
title: "EPID 674: Descriptive Statistics"
subtitle: "Homework 4 Template"
author: "Insert your name here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
format: html
editor: source
---

# Load libraries

```{r}
#| label: load_libraries

library(tidyverse)
library(here)
library(gtsummary)
library(flextable)
```

# Load the dataset

```{r}
#| label: load_dataset

# Load saved NHANES homework dataset from the R project files. This is the dataset we built in homework 2
load(here(("nhanes_homework_dataset.rda")), verbose = TRUE)
```

# 1. Calculate univariate descriptive statistics on a numeric variable

### Calculate descriptive statistics on the cognition variable. For the cognition variable (CFDDS), calculate the mean, standard deviation, minimum, 25th percentile, median, 75th percentile, and maximum values.

```{r}
#| label: univariate_numeric_cognition

# Insert code for problem 1 here

```

Interpret the findings:

# 2. Calculate univariate descriptive statistics for a factor variable

### For the educational attainment variable (education), calculate the number of participants in each category and the percent of the total sample in each category.

```{r}
#| label: univariate_factor_education

# Insert code for problem 2 here

```

Interpret the findings:

# 3. Filter dataset and calculate univariate descriptive statistics

### Filter to participants who have mild cognitive impairment. Calculate the number of participants in each educational attainment category and the percent of the total sample in each category.

```{r}
#| label: filter_univariate

# Insert code for problem 3 here

```

Interpret the findings:

# 4. Calculate bivariate descriptive statistics

### Use the tbl_summary() function to calculate bivariate descriptive statistics on your dataset, split by mild cognitive impairment status (variable name: MCI). Include the following variables as rows in your table (variable names: sex, RIDAGEYR, race_eth, INDFMPIR, education, LBXCOT, LBXBPB, LBXBCD).

```{r}
#| label: bivariate_descriptives

# Insert code for problem 4 here

```

Interpret the findings:
