---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Homework 3B"
author: "Insert Your Name Here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
---


```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(foreign)
library(epiDisplay)
library(stats)
library(Hmisc)
library(gmodels)

directory <- "/cloud/project/"
```

## Using the dataset “bpa.rda”, answer the following:

Type-2 diabetes mellitus (T2DM) can be defined as hemoglobin A1c ≥ 6.5% OR use of diabetes medication. Note: participants with high A1c and taking medication should also be considered T2DM cases.
Create a binary variable for T2DM using a1c and dmmed.

# Problem 1
## How many cases (N) of T2DM are in this population? What is the frequency (percentage) of T2DM in the population?


```{r problem 1}
# Insert code for problem 1 here
```
# Interpret your answer for problem 1 here


#Problem 2. 
##Compute the means and SDs of waist circumference and total cholesterol by T2DM and sex.
```{r problem 2}
# Insert code for problem 2 here
```
# Interpret your answer for problem 2 here
